# Sample Chapter: Introduction to Mathematical Modelling

## Foundations {#sec:foundations}

::: {.definition name="Model" #Model}
A mathematical model is a mathematical framework describing a simplified version of reality.
:::

Some examples of mathematical models from Definition \@ref(def:Model) include

- **Population Growth Model**: A model that describes how the size of a population changes over time depending on birth rates and death rates.

- **Supply and Demand Model**: A model in economics that relates the price of a good to its availability (supply) and how much people want it (demand). It predicts that the price will adjust until the amount supplied equals the amount demanded (market equilibrium).

- **Projectile Motion Model**: A physics model for the path of an object thrown into the air.

- **Linear Regression Model**: A statistical model that expresses the relationship between a response variable (like exam score) and one or more explanatory variables (like study hours).



## Exponential Growth and Decay

Exponential models describe processes that grow or decline at rates proportional to their current size. In the case of exponential growth, the model is

$$
\frac{dN}{dt} = rN, \quad r > 0,
$$

which has the solution

$$
N(t) = N_0 e^{rt}.
$$

When the rate $r$ is negative, we obtain exponential decay, where the quantity decreases over time:

$$
N(t) = N_0 e^{rt}, \quad r < 0.
$$

These models are widely used in contexts such as population growth, radioactive decay, and compound interest.


```{r ExpGrowthDecay, echo=FALSE, fig.align='center', fig.cap="Exponential growth (left) and exponential decay (right) with initial value $N_0=10$; rates $r=0.3$ and $r=-0.3$.", fig.show='hold', out.width=c("50%","50%")}
# Exponential model
exp_model <- function(t, r, N0) N0 * exp(r * t)

# Parameters and grid
N0 <- 10
t  <- seq(0, 10, length.out = 200)

# Plot: growth
plot(t, exp_model(t, r = 0.3,  N0 = N0), type = "l", lwd = 2, col = "blue",
     xlab = "Time (t)", ylab = "N(t)", main = "Exponential Growth")
abline(h = N0, lty = 2, col = "gray")

# Plot: decay
plot(t, exp_model(t, r = -0.3, N0 = N0), type = "l", lwd = 2, col = "red",
     xlab = "Time (t)", ylab = "N(t)", main = "Exponential Decay")
abline(h = 0, lty = 2, col = "gray")
```


## Logistic Model

The logistic growth model describes how a population grows over time with limited resources.

The differential equation form is

$$
\frac{dN}{dt} = rN \left(1 - \frac{N}{K}\right),
$$

where  
- $N(t)$ = population size at time $t$  
- $r$ = intrinsic growth rate  
- $K$ = carrying capacity.  

The solution to this differential equation is

$$
N(t) = \frac{K}{1 + \left(\frac{K - N_0}{N_0}\right)e^{-rt}},
$$

where $N_0$ is the initial population size at $t=0$.  

As $t \to \infty$, $N(t) \to K$, meaning the population stabilizes at the carrying capacity.


```{r LogisticExample, fig.cap="An illustration of the logisitc model for a growth rate of 0.5, a carrying capacity of 100 and an initial population size of 5.", echo=FALSE, fig.align='center', warning=FALSE}
# Logistic growth function
logistic <- function(t, r, K, N0) {
  K / (1 + ((K - N0) / N0) * exp(-r * t))
}

# Parameters
r  <- 0.5      # intrinsic growth rate
K  <- 100      # carrying capacity
N0 <- 5        # initial population size

# Time grid
t <- seq(0, 20, length.out = 200)

# Population values
N <- logistic(t, r, K, N0)

# Plot
plot(t, N, type="l", lwd=2, col="steelblue2",
     xlab="Time (t)", ylab="Population size N(t)",
     main="Logistic Growth Model")
abline(h=K, lty=2, col="lightcoral")
legend("bottomright", legend=c("Population", "Carrying capacity (K)"),
       col=c("steelblue2", "lightcoral"), lty=c(1,2), lwd=2, bty="n")
```

